---
title: "Impact of storms on health and economics across the United States"
author: "Torben Griebe"
date: "17. August 2014"
output: html_document
---

Synopsis
--------
This report tries to find the impact of different storm events on population
health and economics across the United States. The storm data provided by the
U.S. National Oceanic and Atmospheric Administration is used to identify:

* the most harmful events with respect to population health and
* the events with the greatest economic consequences. 

The results show, that tornados are by far the most harmful events with respect 
to population health. In contrast the greatest economic consequences are
caused by floods. 

Data loading and processing
---------------------------
The necessary 
[storm data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) 
is obtained from the U.S. National Oceanic and Atmospheric Administration's 
storm database and stored into a subfolder named `data`.
```{r}
file.name <- "data/repdata_data_StormData.csv.bz2"
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
if (!file.exists(file.name)) {
  dir.create("data")
  # OS dependencies
  switch(Sys.info()[['sysname']],
         Windows = {download.file(url, file.name)},
         Darwin = {download.file(url, file.name, method="curl")})
}
```
The events in the downloaded data file start in the year 1950 and end in 
November 2011.

### Reading in the data
In a first step, the raw data obtained in the previous step is read in. 
The data in the uncompressed file is a comma separated file. The first row is a header line. Missing values are simply left out. 
```{r, cache=TRUE}
storm.data <- read.csv(file.name, na.strings="")
```
After reading in the data, the number of observations and variables are obtained
to get a first insight in the data available. 
```{r}
dim(storm.data)
```
As stated, there are 902297 observations of 37 different variables.

### Relevant variables
To get an idea of the relevant variables, the structure of the data is examined
in combination with the National Weather Service 
[Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
(from here on out called "SSD") which was provided as part of the 
peer assessment 2 description of the Coursera "Reproducible Research" course.
```{r}
str(storm.data)
```
As stated in chapter 2.6 of the SDD and in reconciliation with the 
variable names, the variables `INJURIES` and `FATALITIES` together specifiy the
impact on the population health.

Additionally the variables `PROPDMG`, `PROPDMGEXP`, `CROPDMG` and `CROPDMGEXP`
describe the economic consequences (see chapter 2.7 of the SDD)

The last variable needed is named `EVTYPE` as stated in the description of the
peer assessement.

### Cleaning up health and economics data

To investigate the two questions, two separate sub-datasets will be created and
separately examined. But before that, the common variable `EVTYPE` is analyzed.

#### Cleaning up common data
After a first look at the values of `EVTYPE` (the values won't be displayed 
here so the report doesn't get overfilled) it turns out, that the values 
are string formatted. 
Furthermore some of them contains leading or double whitespaces 
or consist of only capital letters. Some of them contain abbreviated words like
"TSTM" for "thunderstorm" or are written in plural or singular.
To simplify the future usage of this variable, unnecessary whitespaces are 
stripped and all letters are converted to lowercase. Additionally, 
nondescriptive characters like slashes or dashes are deleted and the 
abbreviation "tstm" is written out. 
These actions remove some duplicated event types. 
```{r}
num.of.levels = length(levels(storm.data$EVTYPE))
storm.data$EVTYPE = tolower(gsub("[&[:punct:]]", " ", 
                                 storm.data$EVTYPE, perl=TRUE))
storm.data$EVTYPE = gsub("^[[:space:]]*(.*?)[[:space:]]*$", "\\1", 
                         storm.data$EVTYPE, perl=TRUE)
storm.data$EVTYPE = gsub("[[:space:]]{2,}", " ", 
                         storm.data$EVTYPE, perl=TRUE)
storm.data$EVTYPE = factor(gsub("tstm", "thunderstorm", storm.data$EVTYPE))
cat(sprintf("Deleted %d duplicates", 
            num.of.levels - length(levels(storm.data$EVTYPE))))
```
There are many more posibilities to consolidate the event types, e.g.: 

* Correction of typing errors
* Writing out more abbreviations
* Removing plural suffixes
* Contextual merging

But due to shortage of time, this isn't part of this report. But in my opinion,
it is definetly an issue for further researches.

#### Cleaning up health data
```{r}
health.data <- storm.data[, c("EVTYPE", "FATALITIES", "INJURIES")]
names(health.data) <- c("event.type", "fatalities", "injuries")
```
To get an overview of the data, a short summary is displayed.
```{r}
summary(health.data)
```
The variables `fatalities` and `injuries` contain only valid values. There are
no NA-values or negative numbers. 

In this report, fatalities and injuries are combined to one value called 
`health.damage`. 
```{r}
health.data$health.damage <- rowSums(health.data[, c("fatalities", "injuries")])
```
For another more in-depth research, they should be kept 
separate or they should be weighted differently before they are summed up. 
At least in my opinion a fatality is more severe than only an injury.

#### Cleaning up economics data
```{r}
economics.data <- storm.data[, c("EVTYPE", "PROPDMG", "PROPDMGEXP", 
                                 "CROPDMG", "CROPDMGEXP")]
names(economics.data) <- c("event.type", 
                           "property.damage", "property.magnitude.char",
                           "crop.damage", "crop.magnitude.char")
```
As before, a short summary is displayed for an overview of the data.
```{r}
summary(economics.data)
```
First of all the variables `property.damage` and `crop.damage` contain no
NA-values or negative numbers. But the corresponding two variables 
`property.magnitude.char` and `crop.magnitude.char` representing the
magnitude numbers seem to contain some undefined and invalid values. As the SDD
specifies, these values should only contain the characters "K", "M" or "B" 
(see SDD, chapter 2.7).


Therefore a closer look at these to variables is required.
On a first try it is checked, if there is any damage specified with no (i.e. NA)
magnitude number.
```{r}
sum(economics.data$property.damage)
sum(economics.data$property.damage[is.na(economics.data$property.magnitude.char)])
sum(economics.data$crop.damage)
sum(economics.data$crop.damage[is.na(economics.data$crop.magnitude.char)])
```
Overall these observations can be ignored in further calculations 
because of there extremly small ratio.


Now the invalid magnitude numbers are inspected.
```{r}
levels(economics.data$property.magnitude.char)
levels(economics.data$crop.magnitude.char)
```
Some of the allowed values are present both in upper and lower case letters. 
This issue is solved by converting all values to lower case.
```{r}
economics.data$property.magnitude.char <- 
  factor(tolower(economics.data$property.magnitude.char))
economics.data$crop.magnitude.char <- 
  factor(tolower(economics.data$crop.magnitude.char))
```
To get an idea of what to do with the other values, the relative occurrence of
them in respect to the occurences of the valid values is analyzed.
```{r}
mean(!economics.data$property.magnitude.char[
  !is.na(economics.data$property.magnitude.char)] %in% c("b", "k", "m"))
mean(!economics.data$crop.magnitude.char[
  !is.na(economics.data$crop.magnitude.char)] %in% c("b", "k", "m"))
```
The ratio is small enough to ignore the invalid values.


Now, the damage represented by the variable `[...].damage` and the 
alphabetical magnitude variable `[...].magnitude.char` can be converted into 
actual damage values. The calculated damage values are stored
in the corresponding variable `[...].damage` again.
```{r, cache=TRUE}
economics.data$property.damage <- 
  apply(economics.data[, c("property.damage", "property.magnitude.char")],
        1, function(x) {          
          switch(x[2],
                 k={as.numeric(x[1]) * 10^3},
                 m={as.numeric(x[1]) * 10^6},
                 b={as.numeric(x[1]) * 10^9},
                 {as.numeric(x[1])})
          })
economics.data$crop.damage <- 
  apply(economics.data[, c("crop.damage", "crop.magnitude.char")],
        1, function(x) {
          switch(x[2],
                 k={as.numeric(x[1]) * 10^3},
                 m={as.numeric(x[1]) * 10^6},
                 b={as.numeric(x[1]) * 10^9},
                 {as.numeric(x[1])})
          })
```

In a last step, the overall damage values are stored into a new variable
called `economics.damage`.
```{r}
economics.data$economics.damage <- 
  rowSums(economics.data[, c("property.damage", "crop.damage")])
```


### Processing health and economics data
After loading and cleaning up the data, the data can be processed to answer the
question mentioned at the beginning of this report.
Therefore the injuries and fatalities, as well as the economic damage are 
aggregated by the event type.
```{r, cache=TRUE}
health.impact <- aggregate(health.damage ~ event.type, 
                           health.data, sum)
economics.impact <- aggregate(economics.damage ~ event.type, 
                              economics.data, sum)
```

The aggregated data is now ordered decreasing by the amount of damage.
```{r}
health.impact <- health.impact[order(health.impact$health.damage, 
                                     decreasing=TRUE), ]
row.names(health.impact) <- seq_len(nrow(health.impact))
                                    
economics.impact <- economics.impact[order(economics.impact$economics.damage, 
                                           decreasing=TRUE), ]
row.names(economics.impact) <- seq_len(nrow(economics.impact))
```

Results
-------
**Question 1: Across the United States, which types of events are most harmful 
with respect to population health?**

The following table show the 20 most harmful storm events. The first column 
describes the event, the second column contains the amount of "damage". In this
case, damage means injuries and fatalities.
```{r, results="asis"}
library(xtable)
print(xtable(head(health.impact, n=20)), type="html")
```

<br />
<br />

**Question 2: Across the United States, which types of events have the greatest 
economic consequences?**

Below are the top 20 storm events displayed that have the greatest economic
consequences across the United States.
As before, the first column describes the event. The second column states the
amount of economics damage in Dollar.
```{r, results="asis"}
print(xtable(head(economics.impact, n=20)), type="html")
```

<br />
<br />

As the results show, some of the event types are duplicates, differing only in
the plural suffix (e.g. "thunderstorm wind" and "thunderstorm winds"). Others
might be handled as contextual identical (e.g. "excessive heat" 
and "heat wave").
Therefore, the optimizations mentioned in the chapter "Cleaning up common data"
would improve the performance of the results. 

The following graph shows the connection between the top event types regarding
population health harm and economics consequences. This way, you can get an idea
of the event types that might cause great damage to both "damage types".
To keep the graph clear, the legend is not displayed. But both the "Health" and
the "Economics" column are ordered as in the corresponding tables above.
```{r}
library(ggplot2)
plot.data <- data.frame(pos=seq(from=20, to=1), 
                        health.event.type=health.impact$event.type[1:20],
                        economics.event.type=economics.impact$event.type[1:20])
plot.data <- melt(plot.data, id.vars=1)
print(g <- ggplot(data=plot.data, aes(variable, pos, group=value, colour=value)) 
      + geom_point(size=5, pch=20, alpha=0.5) 
      + geom_line()
      + scale_x_discrete(labels=c("Health", "Economics"))
      + scale_y_continuous(breaks=seq(from=1, to=20), 
                           labels=seq(from=20, to=1))
      + theme(legend.position="none"))
```

