---
title: "Impact of storms on health and economics across the United States"
author: "Torben Griebe"
date: "17. August 2014"
output: html_document
---

Impact of storms on health and economics across the United States
=================================================================

Synopsis
--------

Data loading and processing
---------------------------
The necessary 
[storm data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) 
was obtained from the U.S. National Oceanic and Atmospheric Administration's 
storm database and stored into a subfolder named `data`.
```{r}
file.name <- "data/repdata_data_StormData.csv.bz2"
url <- https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2
if (!file.exists(file.name)) {
  dir.create("data")
  # OS dependencies
  switch(Sys.info()[['sysname']],
         Windows = {download.file(url, file.name)},
         Darwin = {download.file(url, file.name, method="curl")})
}
```
The events in the downloaded data file start in the year 1950 and end in 
November 2011.

### Reading in the data
In a first step, the raw data obtained in the previous step was read in. The 
file was compressed using the format "bz2". The data in the uncompressed file
is a comma separated file. The first row is a header line. Missing values are
simply left out.
```{r, cache=TRUE}
storm.data <- read.csv(file.name, na.strings="")
```
After reading in the data, the number of observations and variables are obtained
to get a first insight in the data available. 
```{r}
dim(storm.data)
```

### Relevant variables
To get an idea of the relevant variables, the structure of the data is examined
in combination with the National Weather Service 
[Storm Data Documentation (SDD)](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) which was provided as part of the peer assessment 2 description of 
the Coursera "Reproducible Research" course.
```{r}
str(storm.data)
```
As stated in chapter 2.6 of the SDD and in reconciliation with the 
variable names, the variables `INJURIES` and `FATALITIES` together specifiy the
impact on the population health.

Additionally the variables `PROPDMG`, `PROPDMGEXP`, `CROPDMG` and `CROPDMGEXP`
describe the economic consequences (see chapter 2.7 of the SDD)

### Processing data

Results
-------

